---
title: "Sensitivity analysis for REDD+ carbon accounting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis for REDD+ carbon accounting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 96
)
```

```{r setup, message = FALSE}
library(mocaredd.dev2)
```

## Why sensitivity analysis?

A REDD+ emission reduction estimate combines many uncertain inputs: areas of land
use change (**activity data**, AD), carbon stocks of each land use, a carbon
fraction, and degradation ratios (together the **emission factors**, EF).
Uncertainty analysis (see `fct_arithmetic_mean2()`) tells you *how uncertain* the
result is. **Sensitivity analysis** answers the complementary question: *which
inputs are responsible for that uncertainty?* This is what tells a country where
to invest the next measurement campaign — improving the input that drives the
variance is the cheapest way to shrink the confidence interval of the reported
emission reduction.

## The emission model

For each land use transition, emissions are activity data times an emission factor:

$$
E \;=\; AD \times EF \times \tfrac{44}{12}, \qquad EF = C_\text{initial} - C_\text{final}
$$

where each carbon stock $C$ is built from carbon pools (optionally through a carbon
fraction and, for degraded land, a degradation ratio applied to the intact stock).
Reference and monitoring emission *levels* are period aggregates of these
transitions, and the **emission reduction** is their difference:

$$
ER = E_\text{REF} - E_\text{MON}.
$$

Because a degraded land use reuses the intact carbon stock, and a global carbon
fraction is shared across land uses, some inputs feed several parts of the model
at once. A correct sensitivity analysis must respect those shared inputs.

## Variance-based sensitivity analysis

The reference framework for global sensitivity analysis is the **variance-based
(Sobol) decomposition** (Saltelli et al. 2008). If the output $Y$ is a function of
independent inputs $X_1,\dots,X_k$, its variance decomposes into contributions of
each input and of their interactions:

$$
\operatorname{Var}(Y) = \sum_i V_i + \sum_{i<j} V_{ij} + \dots
$$

Two indices summarise this:

- the **first-order index** $S_i = V_i / \operatorname{Var}(Y)$, the share of
  variance explained by $X_i$ *alone*;
- the **total index** $S_{T_i}$, which also includes every interaction involving
  $X_i$.

Sobol indices are usually estimated by Monte Carlo with Saltelli sampling, at a
cost of $N\,(k+2)$ model runs. That is the most general approach and it captures
interactions, but it is stochastic and comparatively expensive.

### The approach used in `mocaredd`

The REDD+ emission model is a sum of products (`AD × EF`), which is close to
linear in each input over its uncertainty range. For such a model the first-order
Sobol indices can be obtained **analytically** from a first-order (delta-method)
expansion. For independent inputs,

$$
\operatorname{Var}(Y) \;\approx\; \sum_i \Big(\tfrac{\partial Y}{\partial X_i}\Big)^2 \sigma_i^2,
\qquad
S_i \;\approx\; \frac{\big(\partial Y/\partial X_i\big)^2 \sigma_i^2}{\operatorname{Var}(Y)} .
$$

`fct_sensitivity2()` computes the partial derivatives numerically over the same
independent inputs used by the Monte Carlo simulation (carbon pools namespaced by
land use, the carbon fraction, degradation ratios and per-transition activity
data), so shared inputs propagate consistently — perturbing a forest's biomass
changes both its intact stock and any degraded stock derived from it. The
per-input contributions are then summed into REDD+-meaningful **input groups**:

- **Activity data** — transition areas,
- **Carbon stock** — the carbon pools of each land use,
- **Carbon fraction** — the biomass-to-carbon conversion,
- **Degradation ratio** — the ratios defining degraded stocks.

This is deterministic, essentially free to compute, and consistent with the IPCC
Tier 1 uncertainty already reported by the package. Its limitation is that, being
first-order, it does not resolve interaction terms; when interactions are expected
to be large, a full Sobol analysis is the appropriate alternative (sketched at the
end of this article).

## Worked example

```{r run}
path    <- system.file("extdata/mocaredd-templatev2-simple.xlsx", package = "mocaredd.dev2")
checked <- fct_checkinput(.path = path)

sa <- fct_sensitivity2(.checked_data = checked)
```

### Which inputs drive the uncertainty?

The variance of each reference / monitoring level (`REF`, `E-MONx`) and of each
emission reduction (`ER-MONx`) is decomposed into the four input groups. The
contributions sum to 100% for each output.

```{r variance-table}
sa$variance
```

```{r variance-plot}
sa$gg_variance
```

Reading the chart: a bar dominated by *Carbon stock* or *Degradation ratio* means
the emission factors drive the uncertainty and better field measurements would
help most; a bar dominated by *Activity data* points instead to the land-use
change mapping.

### Splitting emissions by REDD+ activity

When both deforestation (DF) and degradation (DG) are reported, emissions are
split by activity, each with its own uncertainty ($U\% = z \cdot se / |mean| \cdot 100$):

```{r split-table}
sa$split
```

```{r split-plot}
sa$gg_split
```

This shows not only how much each activity contributes to the total emissions but
also how well each is constrained — degradation is often the more uncertain term.

## The Sobol alternative

When interactions between inputs matter, estimate variance-based indices by Monte
Carlo. The model here is already available as a function of its inputs, so the
inputs can be resampled from their distributions and passed through the same
calculation chain. Conceptually (not run):

```{r sobol-sketch, eval = FALSE}
# install.packages("sensitivity")
library(sensitivity)

# model(X): a matrix of input draws -> vector of ER estimates
# built from the same primitives fct_sensitivity2() perturbs
X1 <- sample_inputs(n = 10000)   # matrix, one column per input
X2 <- sample_inputs(n = 10000)

si <- sobolSalt(model = redd_model, X1, X2, scheme = "A", nboot = 100)
plot(si)                          # first-order and total-order indices
```

`sobolSalt()` returns both first-order and total indices; comparing them reveals
how much variance lives in interactions. For the mostly multiplicative REDD+ model
the analytical first-order decomposition above is usually within a few percent of
the Monte Carlo first-order indices, which is why it is the default here.

## References

- IPCC (2006, 2019 refinement) *Guidelines for National Greenhouse Gas Inventories*,
  Volume 1, Chapter 3 — Uncertainties.
- Saltelli, A. et al. (2008) *Global Sensitivity Analysis: The Primer*. Wiley.
